Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-05-06 VegOzone.tex
Restructure Section 2 - not spell checked

The described methods for risk assessment rely on accurate representation of plant physiological response to key climate variables and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, Boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as subarctic climates has not been performed and relies on assuming that less extreme Boreal parameterization will be applicable. It is important to understand the implications of using these more generic parameterizations for more extreme climate regions. We explore how risk estimates might be over or underestimated using generic vs bespoke subarctic parameterization. We also explore the role of \chem{[O_3]} variability and changing climate variables to understand how this risk might alter in the future with the advent of climate change. To this end, conditions at Svanhovd for the growing seasons 2018/19 shall serve as a reference for present and probable future conditions in northern Fennoscandia.
This allows an understanding of how the threat from ozone may change in the future and which are the key aspects of plant physiology that might determine the potential risk. Hence, we determine which parameters require special attention to define accurately for pollution impact assessment and more generally for describing gas exchange response to a changing climate that will influence biochemical cycling. This might also be one key to solve long-standing issues of earth system models (ESM) which often lack PFTs specifically representing subarctic conditions in their land surface models \citep{GMD:Poulter2015,JAMES:Lawrence2019}.\\

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive multi-annual means (referred to as climatologies) for ground-level ozone concentration and environmental key variables (temperature, precipitation, and global irradiance) based on in situ observations and evaluate the deviation of 2018/19 from the norm (referred to as anomalies). We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for the observed ozone damage in 2018 (Section~\ref{sec:do3se}). In this context, we introduce the $\mathrm{DO_3SE}$ model equations and derive bespoke parameterizations for subarctic species to assess systematic uncertainties arising from the use of default generic parameterizations established from less climatically extreme conditions. In Section~\ref{sec:conc}, we will summarize our results and discuss how this research can support improvements in future assessments of pollution risk and more generally bio-geochemical cycling.

%%
\section{Data}
\label{sec:data}

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on comparing the anomalous warm, bright and dry 2018 growing season with the more climatically average 2019 season to understand the role these environmental conditions may have played in influencing the sensitivity of vegetation to ozone damage. The NIBIO operated Environment Centre Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley serves as the main reference for conditions in northern Fennoscandia. Throughout the paper, we will use Svanhovd synonymously for the atmospheric monitoring site at Svanvik/Pasvik. The available meteorological monitoring data are described in Section~\ref{subsec:atmo_svanvik}. Other data used for cross-evaluation of the ozone monitoring data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We, in particular, compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use these to evaluate reanalysis products for ground-level ozone (Section~\ref{subsubsec:ozone_rea}).

\subsection{NIBIO Environment Centre Svanhovd and atmospheric monitoring}
\label{subsec:atmo_svanvik}
The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The center focuses on the research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths. The locations of the atmospheric monitoring site and the ozone garden are marked in aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). Svanhovd is located south of the settlement of Svanvik, where the Pasvik river, which is the border to Russia, lies only a few hundred meters to the east.\\

@@

An ozone garden was installed over the course of our project at the NIBIO Environment Centre Svanhovd and was likely the northern-most ozone garden to date {\bf TODO: Reference?}. An ozone garden consists of selected plant species that are more sensitive to ozone and display visible injuries. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

To relate observed damage on ozone sensitive plants to local ozone concentrations, an ozone monitor had been installed during the growing seasons 2018/19 in addition to existing monitors of atmospheric key variables. All variables were obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ}. Our ozone monitoring data has also been added to the EBAS database operated by \citet{NILU_EBAS} and is thus openly accessible.
Long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.
All relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. Ozone concentrations \chem{[O_3]} measured at $2\,\unit{m}$ height above ground are averaged hourly. The hatched areas mark times when no ozone data were recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

 As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above-mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. We present a reconstruction concerning the missing data in Section~\ref{subsec:ozone_reco}. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.
Hourly averaged $2\,\unit{m}$ temperatures above $20\,\unit{^\circ C}$ occurred more regularly in July 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, spring temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April.
More rain events with accumulated daily precipitation ($\sum_d \mathrm{Precip}$) above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c).
Qualitatively, global irradiance $Q_0$ (Fig.~\ref{fig:data_svanvik_2018_2019}d) was higher in spring (May) and summer (July) 2018 compared to 2019, while June 2019 showed higher irradiance than 2018. A high global irradiance in most cases is the result of a low cloud fraction. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached in June as naturally expected.

\begin{figure*}[t]
  %\includegraphics[width=12cm]{data_svanvik_2018_2019}
  \includegraphics[width=12cm]{fig02}
  \caption{Observational data from atmospheric monitoring at Svanhovd in 2018/19. The hatched areas indicate periods without ozone monitoring data. (a) Hourly averaged ozone concentration; (b) hourly averaged temperature; (c) daily accumulated precipitation; (d) hourly averaged global irradiance.}
  \label{fig:data_svanvik_2018_2019}
\end{figure*}

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the higher \chem{[O_3]} in 2018 in relation to regional averages and trends, we derive a ground-level ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated continuously by NILU at Svanhovd, but no ozone measurements have been conducted until the start of our project. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we cross-calibrate our data with long-term ozone observations at other sites in northern Fennoscandia. Concerning ozone, available observational data are sparse in this region. An overview of the geographical locations of conducted ozone monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data are described in Section~\ref{subsubsec:ebas}. 
As an additional complement to observational data which are not available at all times and spatially sparse, we also evaluate \chem{[O_3]} derived from a variety of reanalysis products (Section~\ref{subsubsec:ozone_rea}).

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{station_map_fennoscandia}
  \includegraphics[width=8.3cm]{fig03}
  \caption{Cap of the North. Locations of past and present (end of 2018) ozone observation sites in northern Fennoscandia were used in this study. For more details see Tab.~\ref{tab:ebas_obs}. The same color coding for each station as displayed herein will be used in all the following figures.}
  \label{fig:station_map_fennoscandia}
\end{figure}

\subsubsection{Ground-level ozone observations}
\label{subsubsec:ebas}
All ground-level ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. In Table~\ref{tab:ebas_obs}, we give an overview of the exact locations and time range of operational data taking. All stations with available long-term observations are located on higher ground than Svanhovd which will result in relatively higher ground-level ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, at higher resolution the vegetation varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above treeline (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass adhering to the \citet{WMOGuide2018} standards (Fig.\ref{fig:svanhovd_research_station}).

@@

\subsubsection{Ozone reanalysis}
\label{subsubsec:ozone_rea}

Details of reanalysis data sets referred to in this work are listed in Table~\ref{tab:ozone_rea}.

There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone, the Monitoring Atmospheric Composition and Climate (MACC) and the newer Copernicus Atmosphere Monitoring Service reanalysis (CAMSRA) \citep{ACP:Inness2013, ACP:Inness2019}. The temporal as well as spatial resolution of these reanalysis products is rather coarse: 3-hourly and $0.75\,\unit{^\circ}\times0.75\,\unit{^\circ}$ or roughly $29.3\,\unit{km}\times 83.4\unit{km}$ at the location of Svanvik. From Copernicus member institutions, higher resolution surface ozone reanalysis data are available for Europe (CAMS Regional Air Quality (CAMSRAQ) reanalysis ensemble). The CAMSRAQ is based on nine European state-of-the-art numerical air quality models. The ensemble mean is at fairly high spatial and temporal resolution compared to the global reanalysis: $0.1\,\unit{^\circ}\times 0.1\,\unit{^\circ}$ or roughly $3.9\,\unit{km}\times 11.1\,\unit{km}$ and 1-hourly, respectively. The periods covered differ and no data is available before the turn of the millennium. The CAMSRA and CAMSRAQ are available in near real time, but only the former covers a period of sufficient length for climate analysis (2003 -- present). All reanalyses use different but, at that time, the latest versions of the operational weather forecast system (OpenIFS) from ECMWF. They differ substantially in the assimilated observational ozone data. The MACC reanalysis assimilates only satellite-derived tropospheric column ozone, while CAMSRA also includes ozone profiles from satellite retrievals. Most importantly for our study, in situ observations from ozone near-surface station networks are only assimilated in the CAMSRAQ reanalysis ensemble. The quality concerning observations in northern Fennoscandia will be discussed in Section~\ref{subsubsec:clim_ozone}. Due to the found substantial discrepancies between reanalysis data and site observations in northern Fennoscandia, we refrain from using any of the global reanalysis data.

\begin{table}[t]
  \caption{Global/regional ozone reanalysis products used in this study.}
  \label{tab:ozone_rea}
\begin{tabular}{llcclcc}
\tophline
Name & Provider & \multicolumn{2}{c}{Resolution} & Time period & Meteorological forcing & \chem{O_3} assimilation\\
&              & spatial & temporal & & \\
\middlehline
MACC & ECMWF & $0.75^\circ \times 0.75^\circ$ & 3-hourly & 2003 -- 2012 & OPS & satellite $^\triangledown$\\
CAMSRA & ECMWF & $0.75^\circ \times 0.75^\circ$ & 3-hourly & 2003 -- 2012 $^\dagger$ & ERA5 / OPS $^\ddagger$ & satellite $^\blacktriangledown$\\
CAMSRAQ & Copernicus & $0.1^\circ \times 0.1^\circ$ & 1-hourly & 2014 -- 2018 $^\dagger$ & OPS $^\star$ & in situ $^\vartriangle$\\
\bottomhline
\end{tabular}
\belowtable{$^\dagger$ Subset of the data used in this study. The full data set extends into near real-time; $^\ddagger$ ERA5 (2003--2016), OPS (later); $^\star$ EURAD uses WRF for downscaling of operational IFS; $^\triangledown$ MLS, OMI - tropospheric column; $^\blacktriangledown$ SCIAMACHY, MIPAS, MLS, OMI, GOME2, SBUV2 - tropospheric column + profile; $^\vartriangle$ METEO France NRT.} % Table Footnotes
\end{table}

\section{Statistical analysis of environmental key variables}
\label{sec:stats}
To improve our understanding and determine the risk of vegetation from high \chem{[O_3]} in subarctic biomes in a changing climate, we first assess the climate conditions and evaluate the growing seasons 2018/19 concerning anomalies from multi-annual means (climatologies). We derive climatologies for temperature, precipitation, global irradiance, and ozone for Svanvik. The ozone measurements will be put into a climatological as well as regional context.

@@

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we will look at derived ozone climatologies from different perspectives.
The multi-annual means of daily mean ozone, for simplicity here without standard deviations, at observation sites in northern Fennoscandia (refer to Table~\ref{tab:ebas_obs} for details) are shown in Supplement Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}. All climatologies except for Svanvik show peak values of about $48\,\unit{ppb}$ in late April. The spring peak values at Svanvik are lower because of the much lower altitude compared to the other monitoring sites. The annual average ozone concentration \chem{\left<[O_3]\right>} declines throughout the growing season until it reaches its minimum in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. At Svanvik, \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than on sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN). The large spread in the case of Svanvik and Janiskoski is due to the lower number of observations.

@@

The density distribution for Svanvik and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display a similar pattern as the northern Fennoscandia climatology (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b). The spring peak occurs slightly earlier (at the beginning of April). As Svanvik is located at a lower altitude than the other sites, the earlier decline of the ozone peak may point to a stronger influence of vegetation on the removal of ozone. \chem{CO_2} uptake by coniferous trees can occur as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013}. The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during nighttime.

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs_spline}
  \includegraphics[width=8.3cm]{fig06}
  \caption{Ozone climatologies. Fennoscandia combines all available data from Jergul/Karasjok, Esrange, and Pallas. The density distributions are shown together with a multi-annual mean of daily ozone concentrations \chem{[O_3]} (climatology). Splines have been fitted through daily mean \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean. Since \chem{[O_3]} are strongly correlated for sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN) (see Fig.~\ref{fig:density_distribution}), data of these have been averaged together to derive a climatology for northern Fennoscandia. (a) Northern Fennoscandia; (b) Svanvik.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}

@@

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatologies_portrait}
  \includegraphics[width=8.3cm]{fig07}
  \caption{Comparison of climatologies derived from observational data (gray band) with three ozone reanalysis products: (a) MACC reanalysis; (b) CAMSRA; (c) CAMSRAQ reanalysis ensemble mean. The sites, represented by the nearest neighboring grid point of the \chem{[O_3]} reanalysis products, are displayed in different markers and colors. The global products (a), (b) do not reproduce observed ground-level ozone well. The MACC reanalysis (a) is in general too low and displays an erroneous seasonal cycle. The CAMSRA (b) matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of CAMSRAQ reanalysis ensemble reproduces the seasonal cycle in northern Fennoscandia, though the modeled ozone concentrations are slightly too low.}
  \label{fig:ozone_climatologies}
\end{figure}

@@

We have studied the effect of acclimations of biomes to a subarctic climate on phyto-toxic ozone uptake. The comparison between 2018 and 2019 conditions allowed us to consider these acclimations and their influence on vegetation sensitivity to \chem{O_3} in light of future changes of key climate variables as may occur under climate change (e.g. increase of frequency and extent of heatwaves). We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was particularly a result of the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden (in particular in mid--end July).

>>>>

The described methods for risk assessment rely on accurate representation of plant physiological response to key environmental variables and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, Boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as subarctic climates has not been performed and relies on assuming that less extreme Boreal parameterization will be applicable. It is important to understand the implications of using these more generic parameterizations for more extreme climate regions. We explore how risk estimates might be over or underestimated using generic vs bespoke subarctic parameterization. We also explore the role of \chem{[O_3]} variability and changing climate variables to understand how this risk might alter in the future with the advent of climate change. To this end, conditions at Svanhovd for the growing seasons 2018/19 shall serve as a reference for present and probable future conditions in northern Fennoscandia.
This allows an understanding of how the threat from ozone may change in the future and which are the key aspects of plant physiology that might determine the potential risk. Hence, we determine which parameters require special attention to define accurately for pollution impact assessment and more generally for describing gas exchange response to a changing climate that will influence biochemical cycling. This might also be one key to solve long-standing issues of earth system models (ESM) which often lack PFTs specifically representing subarctic conditions in their land surface models \citep{GMD:Poulter2015,JAMES:Lawrence2019}.\\

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive multi-annual means (referred to as climatologies) for ground-level ozone concentration and key environmental variables (temperature, precipitation, and global irradiance) based on in situ observations and evaluate the deviation of 2018/19 from the norm (referred to as anomalies). We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for the observed ozone damage in 2018 (Section~\ref{sec:do3se}). In this context, we introduce the $\mathrm{DO_3SE}$ model equations and derive bespoke parameterizations for subarctic species to assess systematic uncertainties arising from the use of default generic parameterizations established from less climatically extreme conditions. In Section~\ref{sec:conc}, we will summarize our results and discuss how this research can support improvements in future assessments of pollution risk and more generally bio-geochemical cycling.

%%
\section{Data acquisition of ozone and key environmental variables}
\label{sec:data}

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on conditions during the 2018/19 growing season at the NIBIO operated Environment Centre Svanhovd. To this end, we collected ozone and atmospheric monitoring data taken at Svanhovd from different distributors. The main reasons for this spread are varring update frequencies and length of available data sets. In terms of ozone, we also use observational data from monitoring sites in northern Finland and Sweden for cross-evaluation. Svanhovd synonymously refers to the atmospheric monitoring site at Svanvik/Pasvik throughout the paper.

%\subsection{NIBIO Environment Centre Svanhovd and bio-monitoring}
%\label{subsec:atmo_svanvik}
The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The center focuses on the research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths.

To relate observed visible damage on ozone sensitive plants to local ozone concentration (\chem{[O_3]}), both an ozone bio-monitor (referred to as ozone garden) and a conventional ozone monitor using the existing atmospheric monitoring infrastructure had been installed during the 2018/19 growing season at the NIBIO centre. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). Svanhovd is located south of the settlement of Svanvik, where the Pasvik river, which is the border to Russia, lies only a few hundred meters to the east. This ozone garden was likely the northern-most to date {\bf TODO: Reference?}. An ozone garden consists of selected plant species that are very sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed qualitatively visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

All measured atmospheric variables including ozone for the 2018/19 growing season were obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ}. The ozone monitoring data have been added to the EBAS database operated by \citet{NILU_EBAS} and are thus openly accessible. All long-term ozone observation data described in Section~\ref{subsec:ozone_data} are also obtained from EBAS. The long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

@@

\subsection{2018/19 growing season}
\label{subsec:growing_season_1819}

All relevant data for the growing season 2018/19 are shown as timeseries in Fig.~\ref{fig:data_svanvik_2018_2019}. Ozone concentrations measured at $2\,\unit{m}$ height above ground are averaged hourly. The hatched areas mark times when no ozone data were recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above-mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak \chem{[O_3]}. In contrast, \chem{[O_3]} only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019. For further analysis (Section~\ref{sec:stats}) and modeling study with the $\mathrm{DO_3SE}$ model (Section~\ref{sec:do3se}), we reconstructed the missing data. The used gap filling method will be described in detail in a companion paper to be submitted.

Hourly averaged $2\,\unit{m}$ temperatures above $20\,\unit{^\circ C}$ occurred more regularly in July 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, spring temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April.
More rain events with accumulated daily precipitation ($\sum_d \mathrm{Precip}$) above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c).
Qualitatively, global irradiance ($Q_0$) displayed in Fig.~\ref{fig:data_svanvik_2018_2019}d was higher in May and July 2018 compared to 2019, while June 2019 showed higher irradiance than 2018. A high global irradiance in most cases is the result of a low cloud fraction. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached in June as naturally expected.

\begin{figure*}[t]
  %\includegraphics[width=12cm]{data_svanvik_2018_2019}
  \includegraphics[width=12cm]{fig02}
  \caption{Observational data from atmospheric monitoring at Svanhovd in 2018/19. The hatched areas indicate periods without ozone monitoring data. (a) Hourly averaged \chem{[O_3]}; (b) hourly averaged temperature; (c) daily accumulated precipitation; (d) hourly averaged global irradiance.}
  \label{fig:data_svanvik_2018_2019}
\end{figure*}

\subsection{Ozone data}
\label{subsec:ozone_data}
To determine the significance of the higher \chem{[O_3]} in 2018 in relation to regional averages and trends, we derive a ground-level \chem{[O_3]} climatology for Svanvik (Section~\ref{sec:stats}). From 1986--1996, ozone monitoring has been operated continuously by NILU at Svanhovd, but no ozone measurements have been conducted before the 2018/19 growing season. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we cross-calibrate our data with long-term ozone observations at sites in northern Finland and Sweden.

%\subsubsection{Ground-level ozone observations}
%\label{subsubsec:ebas}
\begin{figure}[t]
  %\includegraphics[width=8.3cm]{station_map_fennoscandia}
  \includegraphics[width=8.3cm]{fig03}
  \caption{Cap of the North. Locations of past and present (end of 2018) ozone observation sites in northern Fennoscandia were used in this study. For more details see Tab.~\ref{tab:ebas_obs}. The same color coding for each station as displayed herein will be used in all the following figures.}
  \label{fig:station_map_fennoscandia}
\end{figure}

As indicated above, all ground-level ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. A comprehensive overview over all past and present ozone observation sites in northern Fennoscandia is given in Table~\ref{tab:ebas_obs} and Fig.~\ref{fig:station_map_fennoscandia}, respectively. All stations with available long-term observations are located at higher altitude than Svanhovd which will result in relatively higher ground-level \chem{[O_3]} in comparison. In the absence of local precursors, this is due to the stratification of the atmosphere leading to an increase in ozone abundance with altitude \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, at higher spatial resolution the vegetation varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above treeline (pine forest) and the surrounding vegetation mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass adhering to the \citet{WMOGuide2018} standards (Fig.\ref{fig:svanhovd_research_station}).

@@

\section{Statistical analysis of key environmental variables}
\label{sec:stats}
To improve our understanding and determine the risk of vegetation from high \chem{[O_3]} in subarctic biomes in a changing climate, we first assess the climate conditions and evaluate the growing seasons 2018/19 concerning anomalies from multi-annual means (climatologies). We derive climatologies for temperature, precipitation, global irradiance, and ozone for Svanvik. The ozone measurements will be put into a climatological as well as regional context.

@@

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we will look at derived ozone climatologies from different perspectives.
The multi-annual means of daily mean ozone, for simplicity here without standard deviations, at observation sites in northern Fennoscandia (refer to Table~\ref{tab:ebas_obs} for details) are shown in Supplement Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}. All climatologies except for Svanvik show peak values of about $48\,\unit{ppb}$ in late April. The spring peak values at Svanvik are lower because of the much lower altitude compared to the other monitoring sites. The annual average \chem{[O_3]} \chem{\left<[O_3]\right>} declines throughout the growing season until it reaches its minimum in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. At Svanvik, \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than on sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN). The large spread in the case of Svanvik and Janiskoski is due to the lower number of observations.

@@

The density distribution for Svanvik and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display a similar pattern as the northern Fennoscandia climatology (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b). The spring peak occurs slightly earlier (at the beginning of April). As Svanvik is located at a lower altitude than the other sites, the earlier decline of the ozone peak may point to a stronger influence of vegetation on the removal of ozone. \chem{CO_2} uptake by coniferous trees can occur as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013}. The average \chem{[O_3]} at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during nighttime.

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs_spline}
  \includegraphics[width=8.3cm]{fig06}
  \caption{Ozone climatologies. Fennoscandia combines all available data from Jergul/Karasjok, Esrange, and Pallas. The density distributions are shown together with a multi-annual mean of daily \chem{[O_3]} (climatology). Splines have been fitted through daily mean \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean. Since \chem{[O_3]} are strongly correlated for sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN) (see Fig.~\ref{fig:density_distribution}), data of these have been averaged together to derive a climatology for northern Fennoscandia. (a) Northern Fennoscandia; (b) Svanvik.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}

@@

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatologies_portrait}
  \includegraphics[width=8.3cm]{fig07}
  \caption{Comparison of climatologies derived from observational data (gray band) with three ozone reanalysis products: (a) MACC reanalysis; (b) CAMSRA; (c) CAMSRAQ reanalysis ensemble mean. The sites, represented by the nearest neighboring grid point of the \chem{[O_3]} reanalysis products, are displayed in different markers and colors. The global products (a), (b) do not reproduce observed ground-level ozone well. The MACC reanalysis (a) is in general too low and displays an erroneous seasonal cycle. The CAMSRA (b) matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of CAMSRAQ reanalysis ensemble reproduces the seasonal cycle in northern Fennoscandia, though the modeled \chem{[O_3]}s are slightly too low.}
  \label{fig:ozone_climatologies}
\end{figure}

@@

We have studied the effect of acclimations of biomes to a subarctic climate on phyto-toxic ozone uptake. The comparison between 2018 and 2019 conditions allowed us to consider these acclimations and their influence on vegetation sensitivity to \chem{O_3} in light of future changes of key environmental variables as may occur under climate change (e.g. increase of frequency and extent of heatwaves). We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was particularly a result of the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden (in particular in mid--end July).
